Abstractions for Multi-Sorted Substitutions

### 2024-03-13 paper/paper.tex
Update.

  \subsection{A Structure for Multi-Sorted Syntax}
  \label{sec:syntax:structure}
  % The regularity of the multi-sorted syntax makes it easy to define a
  % structure for syntaxes of arbitrary object-languages, i.e.\ syntaxes
  % with arbitrary amounts of syntactic categories and variables:
  The regularity of the multi-sorted syntax makes it easy to define a
  structure for arbitrary syntaxes, i.e. with arbitrarily many syntactic categories and variable types:
  \LibCode\KSyntax
  The first three fields record the definitions of sorts, terms, and variable introduction.
  The last field records that variable introduction
  \ACode{\AField{`\_}} is injective, which is trivially true for
  constructors. The instantiation for our System F syntax is
  straightforward:
  \AppCode\FSyntaxInst

@@

  A renaming \ACode{S₁ \ADef{→ᵣ} S₂} maps variables from \ACode{S₁} to
  variables from \ACode{S₂}.
  A substitution \ACode{S₁ \ADef{→ₛ} S₂} maps variables from \ACode{S₁} to
  terms with free variables from \ACode{S₂}.

@@

  \begin{figure}
    \centering
    \begin{minipage}[t]{0.5\linewidth}
    \LibCode\KMap
    \LibCode\KAp
    \LibCode\KLiftAlt
    \LibCode\KSingleAlt
    \end{minipage}
    \begin{minipage}[t]{0.4\linewidth}
    \LibCode\KWeakenAlt
    \LibCode\KEq
    \LibCode\KFunExt
    \LibCode\KId
    \LibCode\KIdLift
    \end{minipage}
    \caption{Map Operations}
    \label{fig:map-ops}
  \end{figure}
  Figure~\ref{fig:map-ops} shows the usual operations for renamings and substitutions
  generalized to maps.\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  The type \ACode{S₁ \ADef{→ₖ} S₂} unifies
    renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}.
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  The operation \ACode{ϕ \& x} applies a map to a variable.
  The operation \ACode{ϕ ↑ s} lifts a map under a binder of sort \ACode{s}.
  The operation \ACode{⦅ x/t ⦆} constructs a singleton map that replaces
  \ACode{\ACon{zero}} with \ACode{x/t} and decreases all other variables by one.
  The \ACode{\AField{weaken s}} map increases all variables by one.
  The \ACode{ϕ₁ $\sim$ ϕ₂} type expresses extensional equality of maps.
    For simplicity, we postulate functional extensionality \ACode{$\sim$-ext}.\footnote{
      The actual implementation does \emph{not} use any postulates.
      % The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
  There is an identity map \ACode{id}.
  The lemma \ACode{id↑-id} states that a lifted identity map is again an identity map.

@@

  The operation of applying a map to a term depends on the concrete
  structure of a term, so we define a structure for it:
  \LibCode\KTraversal
  The fields of this structure have the following meaning:
  \begin{itemize}
  \item
    \ACode{t \AField{⋯} ϕ} applies the map \ACode{ϕ} (a renaming or
    substitution) to the term \ACode{t}.
  \item
    \ACode{\AField{⋯-var}} states that applying a map \ACode{ϕ} to a variable
    term \ACode{\AField{`} x}, is the same as applying \ACode{ϕ} to
    the variable \ACode{x}, and then converting the result to a term
    via \ACode{\AField{id/`}}.
  \item
    \ACode{\AField{⋯-id}} states that applying the identity map
    \ACode{id} to a term does not change the term.
  \end{itemize}

>>>>

  \subsection{A Structure for Multi-Sorted Syntax}
  \label{sec:syntax:structure}
  % The regularity of the multi-sorted syntax makes it easy to define a
  % structure for syntaxes of arbitrary object-languages, i.e.\ syntaxes
  % with arbitrary amounts of syntactic categories and variables:
  The regularity of the multi-sorted syntax makes it easy to define a
  structure for arbitrary syntaxes, i.e. with arbitrarily many syntactic categories and variable types:
  \LibCode\KSyntax
  \noindent The first three fields record the definitions of sorts, terms, and variable introduction.
  The last field records that variable introduction
  \ACode{\AField{`\_}} is injective, which is trivially true for
  constructors. The instantiation for our System F syntax is
  straightforward:
  \AppCode\FSyntaxInst

@@

  \noindent A renaming \ACode{S₁ \ADef{→ᵣ} S₂} maps variables from \ACode{S₁} to
  variables from \ACode{S₂}.
  A substitution \ACode{S₁ \ADef{→ₛ} S₂} maps variables from \ACode{S₁} to
  terms with free variables from \ACode{S₂}.

@@

  \begin{figure}
    \centering
    \begin{minipage}[t]{0.5\linewidth}
    \LibCode\KMap
    \LibCode\KAp
    \LibCode\KLiftAlt
    \LibCode\KSingleAlt
    \end{minipage}
    \begin{minipage}[t]{0.4\linewidth}
    \LibCode\KWeakenAlt
    \LibCode\KEq
    \LibCode\KFunExt
    \LibCode\KId
    \LibCode\KIdLift
    \end{minipage}
    \caption{Map Operations}
    \label{fig:map-ops}
  \end{figure}
  Figure~\ref{fig:map-ops} shows the usual operations for renamings and substitutions.\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  The type \ACode{S₁ \ADef{→ₖ} S₂} unifies
    renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}.
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  The operation \ACode{ϕ \& x} applies a map to a variable.
  The operation \ACode{ϕ ↑ s} lifts a map under a binder of sort \ACode{s}.
  The operation \ACode{⦅ x/t ⦆} constructs a singleton map that replaces
  \ACode{\ACon{zero}} with \ACode{x/t} and decreases all other variables by one.
  The \ACode{\AField{weaken s}} map increases all variables by one.
  The \ACode{ϕ₁ $\sim$ ϕ₂} type expresses extensional equality of maps.
    For simplicity, we postulate functional extensionality \ACode{$\sim$-ext}.\footnote{
      The actual implementation does \emph{not} use any postulates.
      % The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
  There is an identity map \ACode{id}.
  The lemma \ACode{id↑-id} states that a lifted identity map is again an identity map.

@@

  The operation of applying a map to a term depends on the concrete
  structure of a term, so we define a structure for it:
  \LibCode\KTraversal
  \noindent The fields of this structure have the following meaning:
  \begin{itemize}
  \item
    \ACode{t \AField{⋯} ϕ} applies the map \ACode{ϕ} (a renaming or
    substitution) to the term \ACode{t}.
  \item
    \ACode{\AField{⋯-var}} states that applying a map \ACode{ϕ} to a variable
    term \ACode{\AField{`} x}, is the same as applying \ACode{ϕ} to
    the variable \ACode{x}, and then converting the result to a term
    via \ACode{\AField{id/`}}.
  \item
    \ACode{\AField{⋯-id}} states that applying the identity map
    \ACode{id} to a term does not change the term.
  \end{itemize}
